
---
name: Lance
title: "Data Visualization Exercises"
output: html_document
---

## Exercise 1: Regional Distribution of a Variable

**Objective:** Explore the distribution of an LFS variable across different regions.

**Task:** Using the `LFS0721-03092022-2015BASED-OPT2-PUF.csv` dataset, create a boxplot to visualize the distribution of an LFS variable (of your choice) across different regions. Be sure to label your axes and give your plot a title. 

**Hints:**
- Import the necessary library for visualization (e.g., `ggplot2`).
- Be cautious about the data type of variables.
- Handle missing values and outliers before generating the plot.

```{r}
library(tidyverse)
library(dplyr)

LFS = read.csv("LFS0721-03092022-2015BASED-OPT2-PUF.csv") 
  
LFS %>% 
  filter(!is.na(C05.Age.as.of.Last.Birthday) & C05.Age.as.of.Last.Birthday < 75) %>% 
  ggplot(.,
         mapping = aes(x = C05.Age.as.of.Last.Birthday)) + 
  geom_boxplot() + facet_wrap(. ~ Region) + xlab("Age") 
```

## Exercise 2: Income Analysis

**Objective:** Analyze the per capita income distribution in urban vs. rural areas.

**Task:** Using the `2021FIESvol1_puf.csv` dataset, create a histogram to visualize the distribution of per capita income for urban and rural areas. Distinguish urban and rural areas using different colors. Label your axes and title your plot.

**Hints:**
- Ensure that the data type of the 'Urban / Rural' variable is suitable for creating a legend.
- Optional: Adjust the binwidth parameter in the histogram as per the data distribution.
- Optional: Take care of any potential outliers or extreme values that might distort the visualization.

```{r}
library(tidyverse)
library(dplyr)

FIES = read.csv("2021FIESvol1_puf.csv")

colnames(FIES)
sapply(FIES,class)

FIES %>% 
  filter(Per.Capita.Income < 500000) %>% 
  mutate(., Urban...Rural = ifelse(
    Urban...Rural == 1,
    "Urban",
    "Rural"
  )) %>% 
  ggplot(.,
         mapping = aes(x = Per.Capita.Income,
                       color = Urban...Rural)) + 
  geom_histogram(binwidth = 5000) +
  facet_wrap(. ~ Urban...Rural) +
  labs(title = "Urban vs Rural Per Capita Income", x = "Per Capita Income", y = "") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
